---
title: "Analyzing NHL Data from 2007-2008 to 2018-2019"
format: 
  html:
    theme: cosmo  # zephyr pulse, sketchy, default, cosmo, vapor etc
author:
  - name: Coby Darer
    email: darerjf@appstate.edu
    affiliation:
      - name: Appalachian State University
        city: Boone
        state: NC  
        url: https://www.appstate.edu/
abstract: >
  This project will explore 5 v 5 data on NHL teams and players from the 2007-2008 season to the 2018-2019 season.
---
## Loading the Necessary Packages
```{r}
#| message: false
library(nhldata)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```
## Introduction

This project will explore some of the 5v5 data on NHL teams from the 2007-2008 season to the 2018-2019 season. Specifically 5v5 scoring over these years. It will look at data on skaters, goalies and teams as a whole. The division alignment has changed within this time frame so the teams will be split up into their current divisions (This is for readability more than comparing separate divisions). 

```{r}
#| code-fold: true
West_conf <- teams %>%
  filter(team %in% c("ANA", "ARI", "CGY", "CHI", "COL", "DAL", "EDM", "LA", "MIN", "NSH", "SJ", "STL", "VAN", "ATL", "WPG", "VGK"))

East_conf <- teams %>%
  filter(team %in% c("BOS", "BUF", "CAR", "CBJ", "DET", "FLA", "MTL", "NJ", "NYI", "NYR", "OTT", "PHI", "PIT", "TB", "TOR", "WSH"))

Atlantic_div <- East_conf %>%
  filter(team %in% c("BOS", "BUF", "DET", "FLA", "MTL", "OTT", "TB", "TOR"))

Metro_div <- East_conf %>%
  filter(team %in% c("CAR", "CBJ", "NJ", "NYI", "NYR", "PHI", "PIT", "WSH"))

Central_div <- West_conf %>%
  filter(team %in% c("ARI", "CHI", "COL", "DAL", "MIN", "NSH", "STL", "ATL", "WPG"))

Pacific_div <- West_conf %>%
  filter(team %in% c("ANA", "CGY", "EDM", "LA", "SJ", "VAN", "VGK"))

```

## Exploring some data `teams`

```{r}
glimpse(teams)
```

There are 28 variables in the teams data set:

Note: Corsi stat in hockey is Shot attempts

1.    `team`: Team name

2.    `season`: Season

3.    `gp`: Games Played
    
4.    `toi` : Time on the ice
    
5.    `cf`: Corsi for
    
6.    `ca`: Corsi against
    
7.    `c_plumin` : Corsi plus/minus (cf-ca)
    
8.    `cf_60`: Corsi for per 60 minutes on the ice
    
9.    `cf_pct`: Corsi for as a percentage of total corsi
    
10.    `ca_60`: Corsi against per 60 minutes on the ice
    
11.    `gf` : Goals scored for a player's team
    
12.    `ga`: Goals scored against player's team
    
13.    `g_plumin`: Goals for - goals against
    
14.    `gf_pct`: Percentage of all goals scored by team
    
15.    `gf_60`: Goals for per 60 minutes on the ice
    
16.    `ga_60`: Goals against per 60 minutes on the ice
    
17.    `xgf`: Expected goals for
    
18.    `xga`: Expected goals against
    
19.    `xg_plumin`: Expected goals for - expected goals against
    
20.    `xgf_pct`: Expected goals for as a percentage of total expected goals for
    
21.    `xgf_60`: Expected goals for per 60 minutes on the ice

22.    `xga_60`: Expected goals against per 60 minutes on the ice

23.    `pent`: Penalties taken

24.    `pend`: Penalties drawn

25.    `p_plumin`: Penalties taken - individual penalties drawn

26.    `sh_pct`: Shooting percentage

27.    `sv_pct`: Save percentage

28.    `pdo`: Just Win Baby


Since we are exploring 5v5, let's see how important it is by looking at each teams goals for.
```{r}
#| code-fold: true
#| warning: false
#| message: false
#| label: fig-1
#| fig-cap: "Number of 5v5 goals for each team each season"
Central_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = gf, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Goals For")

Pacific_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = gf, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Goals For")

Atlantic_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = gf, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Goals For")

Metro_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = gf, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Goals For")
```

Now let's look at shot attempts over the years.
```{r}
#| code-fold: true
#| warning: false
#| message: false
#| label: fig-5
#| fig-cap: "Number of 5v5 shot attempts for each team each season"
Central_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = cf/gp, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Shot Attempts")

Pacific_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = cf/gp, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Shot Attempts")

Atlantic_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = cf/gp, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Shot Attempts")

Metro_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = cf/gp, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Shot Attempts")
```

So it looks like there was some fluctuation, but on average teams are scoring slightly more on 5v5 now than they used to.

## Exploring some data on `skaters`

```{r}
glimpse(skaters)
```

There are 48 variables in the skaters data set:

Note: Corsi stat in hockey is Shot attempts

1.    `player`: Player name

2.    `season`: Season

3.    `team`: First team player played for in a given season
    
4.    `team2` : Second team player played for in a given season
    
5.    `team3`: Third team player played for in a given season
    
6.    `position`: Player's first position
    
7.    `position2` : Player's second position
    
8.    `position3`: Player's third position
    
9.    `gp`: Games Played
    
10.    `toi`: Time on the ice
    
11.    `g` : Goals scored
    
12.    `a`: Assists
    
13.    `p`: Points
    
14.    `p1`: Primary points (goals + primary assists)
    
15.    `p_60`: Points per 60 minutes on the ice
    
16.    `p1_60`: Primary points per 60 minutes on the ice
    
17.    `gs`: Game score
    
18.    `gs_60`: Game score per 60 minutes on the ice
    
19.    `cf`: Corsi for (shot attempts by player's team while player is on the ice)
    
20.    `ca`: Corsi against (shot attempts by opposing team will player is on the ice)
    
21.    `c_plumin`: Corsi plus/minus (cf-ca)

22.    `cf_pct`: Corsi percentage

23.    `rel_cf_pct`: Relative corsi percentage

24.    `gf`: Goals scored for a player's team

25.    `ga`: Goals scored against player's team

26.    `g_plumin`: Shooting Goals for - goals against

27.    `gf_pct`: Save Percentage of all goals scored by player's team

28.    `rel_gf_pct`: Relative goals for percentage

29.    `xgf`: Expected goals for

30.    `xga`: Expected goals against

31.    `xg_plumin`: Expected goals for - expected goals against

32.    `xgf_pct`: Expected goals for as a percentage of a team's total expected goals for

33.    `rel_xgf_pct`: Relative expected goals for as a percentage of a team's total expected goals for

34.    `ipent`: Individual penalties taken

35.    `ipend`: Individual penalties drawn

36.    `ip_plumin`: Individual penalties taken - individual penalties drawn

37.    `icf`: Individual corsi for

38.    `icf_60`: Individual corsi for per 60 minutes on the ice

39.    `ixgf`: Individual expected goals for

40.    `ixgf_60`: Individual expected goals for per 60 minutes on the ice

41.    `ish_pct`: Individual shooting percentage

42.    `pdo`: Just Win Baby

43.    `zsr`: Zone start ratio

44.    `toi_pct`: Percentage of team's total time on ice played by player

45.    `toi_pct_qot`: Percentage of team's total time on ice played by player, weighted by quality of teammates

46.    `cf_pct_qot`: Corsi for percentage weighted by quality of player's teammates

47.    `toi_pct_qoc`: Percentage of team's total time on ice played by player, weighted by quality of opponents

48.    `cf_pct_qoc`: Corsi for percentage weighted by quality of player's opponents

It would be nice to see if the individual player stats reflect this slight increase in production.
```{r}
#| code-fold: true
#| label: fig-9
#| fig-cap: "Number of players with 50 or more 5v5 points"
skaters %>%
  filter(season != "2012-2013") %>%
  filter(p >= 50) %>%
  ggplot(aes(x = season)) +
    geom_bar(color = "black", fill = "lightblue") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Count")
```

```{r}
#| code-fold: true
#| label: fig-10
#| fig-cap: "Number of players with at least 20% shooting percentage"
skaters %>%
  filter(season != "2012-2013") %>%
  filter(ish_pct >= 20) %>%
  ggplot(aes(x = season)) +
  geom_bar(color = "black", fill = "lightblue") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Count")
```

```{r}
#| code-fold: true
#| label: fig-11
#| fig-cap: "Number of players with at least 10% shooting percentage"

skaters %>%
  filter(season != "2012-2013") %>%
  filter(ish_pct >= 10) %>%
  ggplot(aes(x = season)) +
  geom_bar(color = "black", fill = "lightblue") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Count")
```

Interesting, the amount of players with at least 50 points seems to be trending up in the last couple of years but is not more than some of the years in the past. Also shooting percentage has stayed relatively consistent, but at 20% there were some spikes.

## Exploring some data on `goalies`

```{r}
glimpse(goalies)
```

There are 16 variables in the teams data set:

1.    `player`: Player name

2.    `season`: Season

3.    `team`: First team player played for in a given season
    
4.    `team2` : Second team player played for in a given season
    
5.    `team3`: Third team player played for in a given season
    
6.    `gp`: Games Played
    
7.    `toi` : Time on the ice
    
8.    `sa`: Shots against
    
9.    `ga`: Goals against
    
10.    `sv_pct`: Save percentage
    
11.    `xsv_pct` : Expected save percentage
    
12.    `dsv_pct`: Dangerous save percentage
    
13.    `ldsv_pct`: Low danger save percentage
    
14.    `mdsv_pc`: Medium danger save percentage
    
15.    `hdsv_pct`: High danger save percentage
    
16.    `gsaa`: Goals saved above average

How have the goalies been affected over the years?
```{r}
#| code-fold: true
#| label: fig-12
#| fig-cap: "Number of goalies with at least 45 games played and a 91% save percentage"
goalies %>%
  filter(season != "2012-2013") %>%
  filter(sv_pct >= 91.00) %>%
  filter(gp >= 45) %>%
  ggplot(aes(x = season)) + 
  geom_bar(color = "black", fill = "pink") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Count")
```

```{r}
#| code-fold: true
#| label: fig-13
#| fig-cap: "Number of who goalies faced an average of at least 20 shots a game"
goalies <- goalies %>%
  mutate(avg_sa_gp = sa/gp)

filtered_goalies <- goalies %>%
  filter(season != "2012-2013") %>%
  filter(gp > 5) %>%
  filter(avg_sa_gp >= 20)

ggplot(filtered_goalies, aes(x = season)) + 
  geom_bar(color = "black", fill = "pink") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Count")
```

## Looking at penalties

```{r}
#| code-fold: true
#| warning: false
#| message: false
#| label: fig-14
#| fig-cap: "Number of penalties taken by each team"
Central_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = pent, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalties Taken")

Pacific_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = pent, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalties Taken")

Atlantic_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = pent, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalties Taken")

Metro_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = pent, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalties Taken")
```

```{r}
#| code-fold: true
#| warning: false
#| message: false
#| label: fig-18
#| fig-cap: "Net penalties by each team"
Central_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = p_plumin, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalty +-")

Pacific_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = p_plumin, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalty +-")

Atlantic_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = p_plumin, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalty +-")

Metro_div %>%
  filter(season != "2012-2013") %>%
  ggplot(aes(x = season, y = p_plumin, color = team)) + 
    geom_point() +
    geom_line(aes(group = team)) +
    geom_smooth(aes(group = 1), se = FALSE) + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    labs(x = "Season",
         y = "Penalty +-")
```

